Contracts for unions and intersections are hard for Nickel

### 2021-06-07 paper.tex
Add numbering to all the subsections

\subsection*{Configuration languages}

@@

\subsection*{Nickel}

@@

\subsection*{Contracts}

@@

\subsection*{Structure of this paper}
\unsure{Make this section into a contributions section?}
Section~\ref{sec:intro} introduced the perspective of designing a configuration
language. We presented the concept of contracts and detailed why contracts are
particularly fitted for a configuration language, and more generally for data
validation or program correctness in a dynamically typed language. The rest of
the paper is structured as follows:
\begin{itemize}
    \item Section~\ref{sec:feat-lang} outlines various common features of
        programming languages that may interact in a bad way with union and
        intersection contracts.
    \item Section~\ref{sec:union-inter} introduces unions and intersections and
        lists the possible motivations for adding them to a language.
    \item Section~\ref{sec:issues-sem} studies intrinsic difficulties one may
        encounter when trying to implement union and intersection contracts in
        practice.
    \item Section~\ref{sec:issues-literature} presents the shortcomings found in
        existing attempts to solve said difficulties.
    \item Section~\ref{sec:related-work} provides some related work.
\end{itemize}

@@

\subsection*{User-defined contracts}
\label{sec:flat-contracts}

@@

\subsection*{Purity}
\label{sec:optimizations}
\info{Present inlining and CSE as two major code
optimizations.
Purity and immutability as two important factors.}

@@

\subsection*{Unions}
Unions ($\cup$) are a fundamental building block of programs.
A union holds a value whose
shape is known to be in a fixed set of alternatives, without exposing
which of those alternatives the value belongs to.
For example, a value that could be either a Number or a String can be
modelled using unions as \nickel{Num @| String}.
They correspond logically to disjunctions.

@@

\subsection*{Intersections}

@@

\subsection*{Union contracts as a side-effect}

@@

This single witness property can be justified as follows. Apart from the error
reporting part (although this is the crucial bit in practice!), the current
contract system of Nickel can be implemented purely as a library, requiring only
a \nickel{fail} primitive to abort the execution. In practice, applying a
function contract to \nickel{f} replaces it with a \nickel{f'} that performs the
additional checks. Thus, since the core language is pure (forgetting the
presence of \nickel{fail}, which isn't relevant), the failure of \nickel{f' 5}
must be independent of its environment and of any previous call to \nickel{f'}.

@@

\begin{figure}[h]
\begin{lstlisting}[language=nickel, title=Original]
let f | Num -> (Bool -> Num @| Bool -> Str)
      = fun x y => if y then x else "False"
in (f 1 true, f 1 false)
\end{lstlisting}
\begin{lstlisting}[language=nickel, title=Optimized]
let f | Num -> (Bool -> Num @| Bool -> Str)
      = fun x y => if y then x else "False"
let g = f 1 in
(g true, g false)
\end{lstlisting}
\caption{Equivalent programs, with inlining or CSE applied}
\label{fig:optimized-programs}
\end{figure}

To sum up, the addition of union contracts introduce side-effects in a pure
language. Side-effects have well-known pitfalls:
\begin{itemize}
    \item For the programmer, they are hard to reason about. They prevent local
        reasoning. In our previous examples, removing or adding a function call
        somewhere can toggle a failure in a call at a totally different
        location.
    \item For the interpreter (or compiler), side-effects inhibit many optimizations and
        program transformations.
\end{itemize}

@@

\subsection*{Intersection with user-defined contracts}
\label{sec:flat-and-inter}

@@

\subsection*{Challenging the semantics}

@@

\subsection*{Keil and Thiemann}
\label{sec:keil-thiemann}
\unsure{Subsection title?}
Keil and Thiemann noticed that, since union contracts should share information between
different application contexts, while intersection contracts should behave
independently between each application context, the system becomes much
simpler if every contract gets transformed into a union of intersections
of other contracts, by using the common De Morgan laws,
and then eagerly opening up unions, but postponing intersection
contracts as much as possible.
\todo{This assumes the difficulty has been previously explained}

@@

\subsection*{Williams, Morris, and Wadler}
\label{sec:will-morr-wadl}
\unsure{Subsection title?}

@@

\subsection*{Racket}
\label{sec:racket}

@@

\subsubsection*{case->}

>>>>

\subsection{Configuration languages}

@@

\subsection{Nickel}

@@

\subsection{Contracts}

@@

\subsection{Structure of this paper}
\unsure{Make this section into a contributions section?}
Section~\ref{sec:intro} introduced the perspective of designing a configuration
language. We presented the concept of contracts and detailed why contracts are
particularly fitted for a configuration language, and more generally for data
validation or program correctness in a dynamically typed language. The rest of
the paper is structured as follows:
\begin{itemize}
    \item Section~\ref{sec:feat-lang} outlines various common features of
        programming languages that may interact in a bad way with union and
        intersection contracts.
    \item Section~\ref{sec:union-inter} introduces unions and intersections and
        lists the possible motivations for adding them to a language.
    \item Section~\ref{sec:issues-sem} studies intrinsic difficulties one may
        encounter when trying to implement union and intersection contracts in
        practice.
    \item Section~\ref{sec:issues-literature} presents the shortcomings found in
        existing attempts to solve said difficulties.
    \item Section~\ref{sec:related-work} provides some related work.
\end{itemize}

@@

\subsection{User-defined contracts}
\label{sec:flat-contracts}

@@

\subsection{Purity}
\label{sec:optimizations}
\info{Present inlining and CSE as two major code
optimizations.
Purity and immutability as two important factors.}

@@

\subsection{Unions}
Unions ($\cup$) are a fundamental building block of programs.
A union holds a value whose
shape is known to be in a fixed set of alternatives, without exposing
which of those alternatives the value belongs to.
For example, a value that could be either a Number or a String can be
modelled using unions as \nickel{Num @| String}.
They correspond logically to disjunctions.

@@

\subsection{Intersections}

@@

\subsection{Union contracts as a side-effect}

@@

This single witness property can be justified as follows. Apart from the error
reporting part (although this is the crucial bit in practice!), the current
contract system of Nickel can be implemented purely as a library, requiring only
a \nickel{fail} primitive to abort the execution. In practice, applying a
function contract to \nickel{f} replaces it with a \nickel{f'} that performs the
additional checks. Thus, since the core language is pure (albeit
partial, if only because \nickel{fail}), the failure of \nickel{f' 5}
must be independent of its environment and of any previous call to \nickel{f'}.

@@

\begin{figure}[h]
\begin{lstlisting}[language=nickel, title=Original]
let f | Num -> (Bool -> Num @| Bool -> Str)
      = fun x y => if y then x else "False"
in (f 1 true, f 1 false)
\end{lstlisting}
\begin{lstlisting}[language=nickel, title=Optimized]
let f | Num -> (Bool -> Num @| Bool -> MStr)
      = fun x y => if y then x else "False"
let g = f 1 in
(g true, g false)
\end{lstlisting}
\caption{Equivalent programs, with inlining or CSE applied}
\label{fig:optimized-programs}
\end{figure}

\unsure{These arguments should go in the purity section}
To sum up, the addition of union contracts introduce side-effects in a pure
language. Side-effects have well-known pitfalls:
\begin{itemize}
    \item For the programmer, they are hard to reason about. They prevent local
        reasoning. In our previous examples, removing or adding a function call
        somewhere can toggle a failure in a call at a totally different
        location.
    \item For the interpreter (or compiler), side-effects inhibit many optimizations and
        program transformations.
\end{itemize}

@@

\subsection{Intersection with user-defined contracts}
\label{sec:flat-and-inter}

@@

\subsection{Challenging the semantics}

@@

\subsection{Keil and Thiemann}
\label{sec:keil-thiemann}
\unsure{Subsection title?}
Keil and Thiemann noticed that, since union contracts should share information between
different application contexts, while intersection contracts should behave
independently between each application context, the system becomes much
simpler if every contract gets transformed into a union of intersections
of other contracts, by using the common De Morgan laws,
and then eagerly opening up unions, but postponing intersection
contracts as much as possible.
\todo{This assumes the difficulty has been previously explained}

@@

\subsection{Williams, Morris, and Wadler}
\label{sec:will-morr-wadl}
\unsure{Subsection title?}

@@

\subsection{Racket}
\label{sec:racket}

@@

\subsubsection{case->}
